Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-10 paper.Rnw
MMC changes

  \item \expfepxval{} is a similar cross-validation to explore \mb{}
  segmentations on the brain images of young, first episode psychosis patients. 
  In addition, \mb{} segmentations with two different atlas segmentation
  protocols are compared to automated segmentations by the \fsl and
  \freesurfer algorithms.  The results of this experiment combined with the
  previous experiment establishes parameter settings that do not overfit to the
  neuroanatomical features of a specific patient cohort.

  \item \expadni{} bridges \mb{} with the existing segmentation literature by
  comparing \mb{} whole hippocampus segmentations with those of several well-known
  automated methods (\freesurfer, \fsl, \maper, \snt) on the entire
  \adnidataset{} image dataset consisting of \Sexpr{length(means.complete$RID)}
  brain images of subjects diagnosed as cognitively normal, having mild cognitive
  impairment, or Alzheimer's disease.

@@

ggplot(fep.volumes, aes(x= mean, y = diff)) + 
  facet_wrap( ~ Method, nrow=2, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  scale_shape_discrete(solid=F,legend=F) + 
  geom_point(size=2,mapping=aes(shape=1)) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean manual and computed volume ' (mm^3))) + 
  ylab(expression('manual - computed volume ' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

@@

Semi-automated segmentations of the left and right whole hippocampi are made
available with a subset of ADNI images \citep{Hsu2002}. These labels have been
generated using the \snt{} tool from Medtronic Surgical Navigation Technologies,
Louisville, CO (see Supplementary Materials for detailed discussion of the
segmentation process).

\maper{} -- a \multiatlas segmentation tool , and \freesurfer{} hippocampal volumes for a subset of images were also
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

@@

One author (MP) performed visual quality inspection for \mb{} and \fsl{}
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass). Only images meeting the conditions of having
segmentations from all methods (\snt{}, \maper{}, \freesurfer{}, \fsl{}, and
\mb{}) and also passing similar quality control inspection were included in the
analysis.

@@

We found a close relationship in total bilateral hippocampal volume between all
methods and the \snt{} semi-automated label volumes (Figure
\ref{fig:ADNI-volumes-plot}). Volumes are correlated with Pearson $r > 0.78$
for all methods across disease categories. Within disease categories (Figure
\ref{fig:ADNI-volumes-boxplot}), \mb{} is consistently well correlated to manual
volumes (Pearson $r > 0.85$), but appears to slightly over-estimate the volume
of the AD hippocampus.

@@

%%%%%%%%%
% ADNI Bland-Altman plots
%%%%%%%%%
<<fig:ADNI-Bland-Altman,dependson='load',fig.show='hide',fig.height=4,fig.width=10>>=
names(means.complete)[8] = "FreeSurfer"
melted=melt(means.complete, measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")

@@

names(melted)[4] = "Diagnosis"
ggplot(melted, aes(x= mean, y = diff, colour=Diagnosis)) + 
  facet_wrap( ~ method, nrow=1, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of SNT and automated volume ' (mm^3))) + 
  ylab(expression('SNT - automated volume ' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

%%%%%%%%%
% Big figure combining the above three ADNI plots
%%%%%%%%%
\begin{figure}
  \centering
  \begin{subfigure}[t]{0.48\textwidth} 
    \includegraphics[width=\textwidth]{figure/fig:ADNI-volumes-plot}
    \caption{Computed vs. semi-automated (\snt) segmentation volume} 
    \label{fig:ADNI-volumes-plot}
  \end{subfigure}%
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI-volumes-boxplot} 
    \caption{Hippocampal volume by diagnosis group and segmentation method} 
    \label{fig:ADNI-volumes-boxplot}
  \end{subfigure}
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI-Bland-Altman}
    \caption{Bland-Altman plots of computed vs. \snt{} hippocampus volume} 
    \label{fig:ADNI-Bland-Altman}
  \end{subfigure}
  \caption{\textbf{\adnidataset{} dataset segmentation.}
(\ref{fig:ADNI-volumes-plot}) Subject mean hippocampal volume as measured by
  each of the four automated methods (\freesurfer{} (FS), \fsl{},
  \maper{}, \mb{}) versus the semi-automated \snt{} segmentation volumes. Linear
  fit lines and Pearson correlations with \snt{} labels are shown for each method.
(\ref{fig:ADNI-volumes-boxplot}) Mean hippocampal volume by method and disease
  category. AD = Alzheimer's disease, LMCI = late-onset mild cognitive
  impairment, and CN = cognitively normal.
(\ref{fig:ADNI-Bland-Altman}) Bland-Altman plots comparing subject mean
  hippocampal volume as measured in the \adnidataset{} dataset by \snt{}
  segmentation and each automated methods. The overall mean difference in volume,
  and limits of agreement ($\pm 1.96SD$) are shown by dashed horizontal lines.
  Linear fit lines are shown for each diagnosis group. Note, points below the
  mean difference indicate overestimation of the volume with respect to the \snt{}
  volume, and vice versa.}
\end{figure}

@@

\paragraph{Winterburn Atlases}  %%%%%%%%%%
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
subfield segmentations of five in-vivo $0.3mm$-isotropic T1-weighted MR images,
described more fully in \expfepxval, section \ref{exp:fep-xval}.

@@

In this experiment, the Winterburn atlases are resampled to $0.9mm$-isotropic
voxel resolution to simulate standard 3T T1-weighted resolution images. Image
subsampling is performed using trilinear subsampling techniques. In each round
of LOOCV, a single atlas image is selected and treated as a target image to be
segmented by \mb{}. So as to have an odd-sized atlas library, atlas image is
segmented once using each possible triple of atlas images, and corresponding
manual segmentations, from the remaining four unselected atlases. Thus, for
each of the five atlases, a total of ${3 \choose 4}=4$ segmentations are
evaluated, resulting in a combined total of $5\times4=20$ segmentations
evaluated overall. We chose an atlas library with an odd number of images so as
to ensure unbiased label fusion when using majority voting (see Discussion).

@@

@@

Figure \ref{fig:WAxval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subfield across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subfields are shown in Table
\ref{tab:WAxval-perturb-kappa}, along with DSC scores for the resampled atlas
segmentations when perturbed slightly and compared to the originals. We find
that the CA4/DG subfield shows the highest mean DSC score of
\Sexpr{mean(CA4DG)}$\pm$\Sexpr{sd(CA4DG)}, followed by the Subiculum and CA1
subfields having scores of \Sexpr{mean(CA1)}$\pm$\Sexpr{sd(CA1)} and
\Sexpr{mean(Sub)}$\pm$\Sexpr{sd(Sub)}, respectively. Both the CA4/DG and
molecular regions score below 0.5. These scores may seem low but not when taken
in context and compared to existing (semi-)automated methods (see Discussion).
The whole hippocampus is segmented with a mean DSC score
of\Sexpr{mean(Whole)}$\pm$\Sexpr{sd(Whole)}.

@@

ggplot(wa.n.all.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=2) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of resampled and MAGeT-Brain subfield volume ' (mm^3))) + 
  ylab(expression('Difference of manual and MAGeT-Brain subfield volume ' (mm^3))) + 
  #scale_x_continuous(limits=c(0,1000)) + 
  #scale_color_discrete(name="") + 
  scale_shape_discrete(name='Subject') + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

@@

caption="Overlap similarity results for the each of the subfields of the hippocampus. 
Simulated overlap similarity results are also given for resampled labels that were 
translated by one voxel (i.e.: $0.9mm$) in all directions. Values are given as
mean Dice's Similarity Coefficient (DSC) $\\pm$ standard deviation"
latex(wa.perturb.table, file="", size="scriptsize",landscape=FALSE, ctable=F,
      caption=caption, caption.loc = c("top"),
      col.just=rep('c',5),  
      label="tab:WAxval-perturb-kappa", rowname=NULL)
@

@@

In this manuscript we have presented the implementation and validation of the
\mb{} framework -- a methodology that requires very few input atlases in order
to provide accurate and reliable segmentations. Both \expadnixval{} (Section
\ref{exp:adni-xval}) and \expfepxval (Section \ref{exp:fepi-xval}) compare \mb{}
to basic-\multiatlas{} segmentation by characterising the change in segmentation
quality with varying parameter settings (atlas and template library sizes,
registration method, and label fusion method) and differing age and
neuropsychiatric populations. Together, these experiments allow us to choose
optimal \mb{} parameter settings for use in subsequent experiments. \expadni{}
(Section \ref{exp:adni}) demonstrates that across
\Sexpr{length(means.complete$RID)} images from the \adnidataset{} dataset, \mb{}
performs as well as, or better, than other established and popular methods.
Finally, \expsubfield{} (Section \ref{exp:winterburn-xval}) demonstrates the
reliability of \mb{} in producing subfield segmentations which match the
segmentation protocol of the input atlases despite contrast and resolution
limitations in standard T1-weighted image volumes.  All of these experiments
together demonstrate that \mb's algorithmic performance is not dependent on a
single definition of the hippocampus but is effective with differing hippocampal
definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}, across image types,
and subject populations.

The core claim the \mb{} method is based on -- that we can meaningfully bootstrap
a template library from a small set of labelled atlas images -- is validated in
the cross validation conducted in \expadnixval{} (and the replication in in
\expfepxval{} and \expadnixvalsnt{}, Supplementary Materials). We find that both
increasing the number of atlases and the number of templates used improves \mb{}
segmentation over and above basic-\multiatlas{} segmentations using the same
number of atlas images. That is, by taking the extra step of generating a
template library using target images, \mb{} is able to improve the overlap
between the automatically generated segmentations and manually generated ``gold
standard'' segmentations. The magnitude of this improvement is greatest with a
small number of atlases, but even with larger atlas libraries we have found that
generating a template library reduces the variability in segmentation precision
(i.e. \mb{} more consistently produces high quality segmentations than does
basic-\multiatlas{} segmentation). These effects do not appear dependant on the
hippocampal segmentation protocol used.

@@

Another concern is the moderate improvement observed in \mb{} in comparison to
\multiatlas{} segmentation when using the same number of atlases. The actual
benefit in using \mb{} is consistency of the labelling regardless of atlas or
template choice, as mentioned above. This is an important consideration that few
have touched on previously. The 10-fold Monte Carlo cross-validation that we
present in \expadnixval{} and \expfepxval{} is amongst one of the most stringent
performed in the \multiatlas{} segmentation literature. To the best of our
knowledge, with the exception of \citep{Wang2011}, other groups using ADNI data
for validation do at most a single round of leave-one-out-validation (Table
\ref{tab:other-methods}). The thoroughness of our validation suggests that our
results are reflective of a true average over the choice of parameter settings
and are independent of atlas or template choice (provided the input atlases are
consistently segmented).

@@

One further complication common to all subfield segmentation evaluation is that,
by its nature, the Dice's Similarity Coefficient score penalizes structures with
high surface area to volume ratios. Therefore subfield DSC scores will generally
be lower than whole hippocampal segmentations. We attempted to put this
effect into perspective by comparing \mb{} subfield segmentation agreement with
the agreement of voxel-shifted manual segmentations (Table
\ref{tab:WAxval-perturb-kappa}). The results of this exercise show conclusively,
despite the very limited number of atlases we had to work with, that \mb{}
subfield segmentations are well within the bounds of error of a $0.3mm$ voxel shift.

Our overlap DSC values demonstrates that we can reliably reproduce segmentations
for the CA1, subiculum, and CA4/dentate subfields (DSC > 0.5). That the CA2/CA3
and molecular layers are less well reproduced (DSC < 0.5) should not be
surprising as these are extremely thin and spatially convoluted regions that
originally required high-resolution MRI for identification and so it is likely
that the extents of these regions are well below the resolution and contrast
offered by standard T1-weighted images.

@@

As when comparing against manual labels in \expadnixval{}, we find similar
behaviour when comparing \mb{} segmentations to \snt{} labels: similarity scores
increase with increasing numbers of atlases and templates, with diminishing
increases in improvement trending towards a plateau (Figure
\ref{fig:ADNI1-xval-k-mean}). As in \expadnixval{}, using \ants{} registration
leads to significantly increased similarity scores, and there is no significant
difference in scores from any of the label fusion methods. Mean DSC score peaks
at \Sexpr{snt.mb.kappa.best} when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion. Compared to
\multiatlas{} segmentations, we find \mb{} segmentations show increasing
improvement with larger atlas and template libraries when using more than 9
templates and 5 or fewer atlases (Figure \ref{fig:ADNI1-SNT-xval-k-diff}). Peak
improvement (+\Sexpr{round(snt.mb.diff.best, digits=3)} DSC) is found with a
single atlas and template library of 19 images. In addition to a mean increase
in similarity score over \multiatlas{}-based segmentation, \mb{} also shows more
consistency in similarity scores across all subjects and validation folds
(Figure \ref{ADNI1-SNT-xval-variability}) with a large enough template library.

>>>>

  \item \expfepxval{} is a similar cross-validation to explore \mb{}
  segmentations on the brain images of young, first episode psychosis patients. 
  In addition, \mb{} segmentations with two different atlas segmentation
  protocols are compared to automated segmentations by the \fsl{} and
  \freesurfer{} algorithms.  The results of this experiment combined with the
  previous experiment establishes parameter settings that do not overfit to the
  neuroanatomical features of a specific patient cohort.

  \item \expadni{} bridges \mb{} with the existing segmentation literature by
  comparing \mb{} whole hippocampus segmentations with those of several well-known
  automated methods (\freesurfer{}, \fsl{}, \maper{}, \snt{}) on the entire
  \adnidataset{} image dataset consisting of \Sexpr{length(means.complete$RID)}
  brain images of subjects diagnosed as cognitively normal, having mild cognitive
  impairment, or Alzheimer's disease.

@@

ggplot(fep.volumes, aes(x= mean, y = diff)) + 
  facet_wrap( ~ Method, nrow=2, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=2) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean manual and computed volume ' (mm^3))) + 
  ylab(expression('manual - computed volume ' (mm^3))) + 
  scale_shape_discrete(solid=F,guide='none') + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.position = "none", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

@@

For a subset of the ADNI images, semi-automated segmentations of the left and
right whole hippocampi generated using the \snt{} tool from Medtronic Surgical
Navigation Technologies, Louisville, CO (see Supplementary Materials for
detailed discussion of the segmentation process) are made available
\citep{Hsu2002}. These labels are used as the reference labels in several other
studies of (semi-)automated segmentation methods (see Discussion).  In addition,
ADNI also distributes hippocampal segmentations and volumes determined using
\maper{} \citep{Heckemann2011}, a \multiatlas{} segmentation tool, and the
\freesurfer{} tool (including quality control data, with guidelines described in
\citet{Hartig2010}).

@@

One author (MP) performed visual quality inspection for \mb{} and \fsl{}
segmentations using similar quality control guidelines to those used by
\freesurfer. If either hippocampus was under or over segmented by $10mm$ or
greater in three or more slices then the segmentation did not pass. Only images
meeting the conditions of having segmentations from all methods (\snt{},
\maper{}, \freesurfer{}, \fsl{}, and \mb{}) and also passing quality control
inspection were included in the analysis.

@@

We found a close relationship in total bilateral hippocampal volume between all
methods and the \snt{} semi-automated label volumes (Figure
\ref{fig:ADNI-volumes-plot}). Volumes are correlated with Pearson $r > 0.78$
for all methods across disease categories. Within disease categories (Figure
\ref{fig:ADNI-volumes-boxplot}), \mb{} is consistently well correlated to manual
volumes (Pearson $r > 0.85$), but appears to slightly over-estimate the volume
of the AD hippocampus compared to the \snt{} segmentations.

@@

%%%%%%%%%
% ADNI Bland-Altman plots
%%%%%%%%%
<<fig:ADNI-Bland-Altman,dependson='load',fig.show='hide',fig.height=6,fig.width=10>>=
names(means.complete)[8] = "FreeSurfer"
melted=melt(means.complete, measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")

@@

names(melted)[4] = "Diagnosis"
ggplot(melted, aes(x= mean, y = diff, colour=Diagnosis)) + 
  facet_wrap( ~ method, nrow=2, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of SNT and automated volume ' (mm^3))) + 
  ylab(expression('SNT - automated volume ' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

%%%%%%%%%
% Big figure combining the above three ADNI plots
%%%%%%%%%
\begin{figure}
  \centering
  \begin{subfigure}[t]{0.48\textwidth} 
    \includegraphics[width=\textwidth]{figure/fig:ADNI-volumes-plot}
    \caption{Computed vs. semi-automated (\snt) segmentation volume} 
    \label{fig:ADNI-volumes-plot}
  \end{subfigure}%
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI-volumes-boxplot} 
    \caption{Hippocampal volume by diagnosis group and segmentation method} 
    \label{fig:ADNI-volumes-boxplot}
  \end{subfigure}
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:ADNI-Bland-Altman}
    \caption{Bland-Altman plots of computed vs. \snt{} hippocampus volume} 
    \label{fig:ADNI-Bland-Altman}
  \end{subfigure}
  \caption{\textbf{\adnidataset{} dataset segmentation.}
(\ref{fig:ADNI-volumes-plot}) Subject mean hippocampal volume as measured by
  each of the four automated methods (\freesurfer{} (FS), \fsl{},
  \maper{}, \mb{}) versus the semi-automated \snt{} segmentation volumes. Linear
  fit lines and Pearson correlations with \snt{} labels are shown for each method.
(\ref{fig:ADNI-volumes-boxplot}) Mean hippocampal volume by method and disease
  category. AD = Alzheimer's disease, LMCI = late-onset mild cognitive
  impairment, and CN = cognitively normal.
(\ref{fig:ADNI-Bland-Altman}) Bland-Altman plots shows the agreement between
  computed and \snt{} hippocampus volume. The overall mean difference in volume,
  and limits of agreement ($\pm 1.96SD$) are shown by dashed horizontal lines.
  Linear fit lines are shown for each diagnosis group. Note, points below the
  mean difference indicate overestimation of the volume with respect to the \snt{}
  volume, and vice versa.}
\end{figure}

@@

@@

In this experiment, the Winterburn atlases (\expfepxval, section
\ref{exp:fep-xval}) are resampled to $0.9mm$-isotropic
voxel resolution to simulate standard 3T T1-weighted resolution images. Image
subsampling is performed using trilinear subsampling techniques. In each round
of LOOCV, a single atlas image is selected and treated as a target image to be
segmented by \mb{}. So as to have an odd-sized atlas library, atlas image is
segmented once using each possible triple of atlas images, and corresponding
manual segmentations, from the remaining four unselected atlases. Thus, for
each of the five atlases, a total of ${3 \choose 4}=4$ segmentations are
evaluated, resulting in a combined total of $5\times4=20$ segmentations
evaluated overall. We chose an atlas library with an odd number of images so as
to ensure unbiased label fusion when using majority voting (see Discussion).

@@

Additionally, by shifting the original manual $0.3mm^3$-isotropic voxel
segmentations by one voxel in the x, y, and z direction and then resampling it
to $0.9mm^3$-isotropic voxels we obtain a simulated manual segmentation having a
small amount of error. We can compare the DSC overlap score of the shifted
labels (relative to the directly resampled labels) with the DSC score of the
\mb{} generated labels in order to evaluate their relevance.

@@

Figure \ref{fig:WAxval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subfield across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subfields are shown in Table
\ref{tab:WAxval-perturb-kappa}. We find that the CA4/DG subfield shows the highest mean DSC score of
\Sexpr{mean(CA4DG)}$\pm$\Sexpr{sd(CA4DG)}, followed by the Subiculum and CA1
subfields having scores of \Sexpr{mean(CA1)}$\pm$\Sexpr{sd(CA1)} and
\Sexpr{mean(Sub)}$\pm$\Sexpr{sd(Sub)}, respectively. Both the CA4/DG and
molecular regions score below 0.5. These scores may seem low but not when taken
in context and compared to existing (semi-)automated methods (see Discussion).
The whole hippocampus is segmented with a mean DSC score
of\Sexpr{mean(Whole)}$\pm$\Sexpr{sd(Whole)}. Table
\ref{tab:WAxval-perturb-kappa} also shows mean DSC overlap scores for simulated
manual segmentations with error introduced by a one-voxel translation in all
directions. For all subfields, \mb{} DSC scores are well within the error bounds
set by these segmentations.

@@

ggplot(wa.n.all.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=2) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of resampled and MAGeT-Brain subfield volume ' (mm^3))) + 
  ylab(expression('Difference of manual and MAGeT-Brain subfield volume ' (mm^3))) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

@@

caption="Overlap similarity results for the each of the subfields of the hippocampus. 
Simulated overlap similarity results are also given for manual labels that were 
translated by one voxel (i.e.: $0.3mm$) in all directions and then resampled.
Values are given as mean Dice's Similarity Coefficient (DSC) $\\pm$ standard
deviation."
latex(wa.perturb.table, file="", size="scriptsize",landscape=FALSE, ctable=F,
      caption=caption, caption.loc = c("top"),
      col.just=rep('c',5),  
      label="tab:WAxval-perturb-kappa", rowname=NULL)
@

@@

In this manuscript we have presented the implementation and validation of the
\mb{} framework -- a methodology that requires very few input atlases in order
to provide accurate and reliable segmentations. Both \expadnixval{} (Section
\ref{exp:adni-xval}) and \expfepxval{} (Section \ref{exp:fep-xval}) compare \mb{}
to basic-\multiatlas{} segmentation by characterising the change in segmentation
quality with varying parameter settings (atlas and template library sizes,
registration method, and label fusion method) and differing age and
neuropsychiatric populations. Together, these experiments allow us to choose
optimal \mb{} parameter settings for use in subsequent experiments. \expadni{}
(Section \ref{exp:adni}) demonstrates that across
\Sexpr{length(means.complete$RID)} images from the \adnidataset{} dataset, \mb{}
performs as well as, or better, than other established and popular methods, and
has a much more conservative proportional bias in segmentation volume.  Finally,
\expsubfield{} (Section \ref{exp:winterburn-xval}) demonstrates the reliability
of \mb{} in producing subfield segmentations which match the segmentation
protocol of the input atlases despite contrast and resolution limitations in
standard T1-weighted image volumes.  All of these experiments together
demonstrate that \mb's algorithmic performance is not dependent on a single
definition of the hippocampus but is effective with differing hippocampal
definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}, across image types,
and subject populations.

The core claim the \mb{} method is based on -- that we can meaningfully bootstrap
a template library from a small set of labelled atlas images -- is validated in
the cross validation conducted in \expadnixval{} (and the replication in
\expfepxval{} and \expadnixvalsnt{}, Supplementary Materials). We find that both
increasing the number of atlases and the number of templates used improves \mb{}
segmentation over and above basic-\multiatlas{} segmentations using the same
number of atlas images. That is, by taking the extra step of generating a
template library using target images, \mb{} is able to improve the overlap
between the automatically generated segmentations and manually generated ``gold
standard'' segmentations. The magnitude of this improvement is greatest with a
small number of atlases, but even with larger atlas libraries we have found that
generating a template library reduces the variability in segmentation precision
(i.e. \mb{} more consistently produces high quality segmentations than does
basic-\multiatlas{} segmentation). These effects do not appear dependant on the
hippocampal segmentation protocol used.

@@

Despite \mb{} achieving segmentation results which are competitive with the rest
of the field (Table \ref{tab:other-methods}), a concern may be raised over the
modest improvement in segmentation agreement observed using \mb{} over
\multiatlas{}, with the same number of atlases (\expadnixval{}). As we have
shown in that same experiment, the benefit in using \mb{} is both an increase in
the overlap agreement and also in the improved consistency of the labelling
regardless of atlas or template choice. Reducing the variability in segmentation
accuracy is an important consideration that few have touched on previously. In
addition, the Monte Carlo cross-validations that we present in \expadnixval{}
and \expfepxval{} are amongst the most stringent performed in the \multiatlas{}
segmentation literature. To the best of our knowledge, with the exception of
\citep{Wang2011}, other groups do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}).  Thus, the
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice.

@@

One further complication common to all subfield segmentation evaluation is that,
by its nature, the Dice's Similarity Coefficient score penalizes structures with
high surface area-to-volume ratios. Therefore subfield DSC scores will generally
be lower than whole hippocampal segmentations. We attempted to put this
effect into perspective by comparing \mb{} subfield segmentation agreement with
the agreement of voxel-shifted manual segmentations (Table
\ref{tab:WAxval-perturb-kappa}). The results of this exercise show conclusively,
despite the very limited number of atlases we had to work with, that \mb{}
subfield segmentations are well within the bounds of error of a $0.3mm^3$ voxel
shift.

Our overlap DSC values demonstrates that we can reliably reproduce segmentations
for the CA1, subiculum, and CA4/dentate subfields ($DSC > 0.5$). That the CA2/CA3
and molecular layers are less well reproduced ($DSC < 0.5$) should not be
surprising as these are extremely thin and spatially convoluted regions that
originally required high-resolution MRI for identification and so it is likely
that the extents of these regions are well below the resolution and contrast
offered by standard T1-weighted images.

@@

As when comparing against manual labels in \expadnixval{}, we find similar
behaviour when comparing \mb{} segmentations to \snt{} labels: similarity scores
increase with increasing numbers of atlases and templates, with diminishing
increases in improvement trending towards a plateau (Figure
\ref{fig:ADNI1-xval-k-mean}). As in \expadnixval{}, using \ants{} registration
leads to significantly increased similarity scores, and there is no significant
difference in scores from any of the label fusion methods. Mean DSC score peaks
at \Sexpr{snt.mb.kappa.best} when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion. Compared to
\multiatlas{} segmentations, we find \mb{} segmentations show increasing
improvement with larger atlas and template libraries when using more than 9
templates and 5 or fewer atlases (Figure \ref{fig:ADNI1-SNT-xval-k-diff}). Peak
improvement (+\Sexpr{round(snt.mb.diff.best, digits=3)} DSC) is found with a
single atlas and template library of 19 images. In addition to a mean increase
in similarity score over \multiatlas{}-based segmentation, \mb{} also shows more
consistency in similarity scores across all subjects and validation folds
(Figure \ref{fig:ADNI1-SNT-xval-variability}) with a large enough template library.
